A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-24 imaging_demo/ms.tex
Add more content under section 5
From 5a9dd6e (master):
=====================

In ms.tex,

(*) Add more content under section 5

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols

@@

Figure~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
after which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Figure~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra from each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

This module is common to the MOFF and FX imaging techniques. Time samples of
electric fields measured by the antenna and digitized by the A/D converter is
Fourier transformed to generate electric field spectra. This step can be
parallelized by antennas as shown in Figure~\ref{fig:f-engine}. The output is
then fed to either MOFF and FX imaging pipelines.

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the antennas
($\mathbf{a}$) to the antenna array grid denoted by the coordinate $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Figure~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel form the output of the correlator in the 
feedback calibration fashion outlined in \citealt{mor11} (illustrated in Figure 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following the 
MOFF algorithm places no restriction on array layout, and accounts for non-identical 
antenna beam patterns. Because only a single correlation is needed for each 
antenna, the computation complexity scales only as $N_{\mathrm{ant}}$.

@@

The shape and extent of these auto-correlations can be estimated from the 
antenna aperture illumination pattern which is a quantity already available 
from the gridding step. Figure~\ref{fig:autocorr_wts_PB} shows the estimated 
weights from antenna auto-correlations in the $uv$-plane (left) and the 
corresponding response in the image plane (right). The latter is simply the 
directional antenna power response.

@@

Figure~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the MOFF image and the corresponding 
synthesized beam. The sky positions of the simulated sources are indicated by 
solid black circles. The reconstructed sky image has the simulated sources at the
expected sky positions in either case. Both algorithms result in images and 
synthesized beams that are well matched with each other. Their fluxes are 
modulated by a multiplicative power pattern corresponding to that of a uniform
square aperture.

@@

Figure~\ref{fig:MOFF-FX-uvwts} shows the cross-correlation weights obtained 
with MOFF imaging (left) and visibility-based imaging (right). The first 
notable difference is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The gridding
process in the former involves rounding the antenna footprint to the nearest 
grid pixel. Depending on the exact location of the center of the antenna 
relative to the grid, the grid pixels that receive contribution from an
antenna may be a pixel narrower along one or both axes relative to that from
another identical antenna but with a different center location relative to the
grid. The resulting auto-correlation footprint will also not necessarily be 
identical. Hence, using a single expected footprint for subtraction will
typically leave some residuals behind as seen in the void region in the 
weights obtained with MOFF imaging.

@@

Figure~\ref{fig:psf-diff} shows the difference between the synthesized beams 
obtained with the two methods. A difference map between the two synthesized 
beams is shown on top. The amplitude of the difference appears to be 
modulated by the directional voltage response of the antenna. At the bottom, 
in radial bins, the rms of the synthesized beam (gray) and the rms of the 
difference map (black) are plotted in percentage units relative to the peak 
(to be read using the axis on the left side of the plot). The antenna power 
pattern (red; to be read using the scale on the right) is plotted for reference.

@@

Figure~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along x-axis and $m$ along y-axis. Due to lack of
calibration the flux scale is arbitrary. Despite the lack of calibration,
only with delay compensation, we see Cyg~A and Cas~A prominently at
$(l,m)\approx (0, 0.125)$, and (0.4,0.55) respectively and Sgr~A faintly at
(-0.5,-0.75), thus validating the functionality of the EPIC package.

@@

We consider antenna array layouts described by three quantities fundamental
to radio interferometry, namely, maximum baseline length, number of antennas,
and the size of each antenna.

% Figure~\ref{fig:parameter-space-bll-nant-instruments} shows the

\subsection{Scaling Relations: MOFF vs. FX}
\subsection{Scaling Up}
\subsection{Case Study}

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding.eps}
  \caption{Current and instruments planned for future in parameter space of
    baseline length and number of antennas with MOFF and FX.}
  \label{fig:parameter-space-bll-nant-instruments}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex multiplies and adds with MOFF and FX.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[flushleft]{threeparttable} % For table notes

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
after which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra from each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

This module is common to the MOFF and FX imaging techniques. Time samples of
electric fields measured by the antenna and digitized by the A/D converter is
Fourier transformed to generate electric field spectra. This step can be
parallelized by antennas as shown in Fig.~\ref{fig:f-engine}. The output is
then fed to either MOFF and FX imaging pipelines.

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the antennas
($\mathbf{a}$) to the antenna array grid denoted by the coordinate $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Fig.~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel form the output of the correlator in the 
feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following the 
MOFF algorithm places no restriction on array layout, and accounts for non-identical 
antenna beam patterns. Because only a single correlation is needed for each 
antenna, the computation complexity scales only as $N_{\mathrm{ant}}$.

@@

The shape and extent of these auto-correlations can be estimated from the 
antenna aperture illumination pattern which is a quantity already available 
from the gridding step. Fig.~\ref{fig:autocorr_wts_PB} shows the estimated 
weights from antenna auto-correlations in the $uv$-plane (left) and the 
corresponding response in the image plane (right). The latter is simply the 
directional antenna power response.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the MOFF image and the corresponding 
synthesized beam. The sky positions of the simulated sources are indicated by 
solid black circles. The reconstructed sky image has the simulated sources at the
expected sky positions in either case. Both algorithms result in images and 
synthesized beams that are well matched with each other. Their fluxes are 
modulated by a multiplicative power pattern corresponding to that of a uniform
square aperture.

@@

Fig.~\ref{fig:MOFF-FX-uvwts} shows the cross-correlation weights obtained 
with MOFF imaging (left) and visibility-based imaging (right). The first 
notable difference is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The gridding
process in the former involves rounding the antenna footprint to the nearest 
grid pixel. Depending on the exact location of the center of the antenna 
relative to the grid, the grid pixels that receive contribution from an
antenna may be a pixel narrower along one or both axes relative to that from
another identical antenna but with a different center location relative to the
grid. The resulting auto-correlation footprint will also not necessarily be 
identical. Hence, using a single expected footprint for subtraction will
typically leave some residuals behind as seen in the void region in the 
weights obtained with MOFF imaging.

@@

Fig.~\ref{fig:psf-diff} shows the difference between the synthesized beams 
obtained with the two methods. A difference map between the two synthesized 
beams is shown on top. The amplitude of the difference appears to be 
modulated by the directional voltage response of the antenna. At the bottom, 
in radial bins, the rms of the synthesized beam (gray) and the rms of the 
difference map (black) are plotted in percentage units relative to the peak 
(to be read using the axis on the left side of the plot). The antenna power 
pattern (red; to be read using the scale on the right) is plotted for reference.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over the entire 2~s of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along x-axis and $m$ along y-axis. Due to lack of
calibration the flux scale is arbitrary. Despite the lack of calibration,
only with delay compensation, we see Cyg~A and Cas~A prominently at
$(l,m)\approx (0, 0.125)$, and (0.4,0.55) respectively and Sgr~A faintly at
(-0.5,-0.75), thus validating the functionality of the EPIC package.

@@

We consider a variety of current and planned radio telescopes. Their antenna 
layouts are summarized in Table~\ref{tab:antenna-layouts}. The size of the 
layout gives the maximum baseline $b_\textrm{max}$. The grid spacing is 
determined by the science goals of the experiment in general. For our 
purpose, we assume a typical requirement that only the field of view of the 
antenna is to be imaged. This sets the grid spacing to be equal to the size 
of the antenna, $A_\textrm{a}$. Hence, 
$N_\textrm{g}\simeq b_\textrm{max}^2/A_\textrm{a}$.

\begin{table}
  \centering
  \caption{Radio telescopes and array layouts.}
  \label{tab:antenna-layouts}
  \begin{threeparttable}
  \begin{tabular}{ccccc} 
    \hline
    Telescope & Core & Number & Antenna & Frequency \\
              & size & of Antennas & size & \\
    (key)     & $b_\textrm{max}$ (in m) & $N_\textrm{a}$ & $A_\textrm{a}$ (in m$^2$) & $f_0$ (in MHz) \\
    \hline
    MWA & 1400 & 112 & 16 & 150 \\
    (MWA-112) \\ \\
    MWA & 1400 & 496 & 16 & 150 \\
    (MWA-496) \\ \\
    LOFAR & 3500 & 24 & & 50 \\
    (LOFAR-C) & & & & \\ \\
    LWA1 & 100 & 256 & 10 & 50 \\
    (LWA1) \\ \\
    LWA-OV\tnote{a} & 200 & 256 & 10 & 50 \\
    (LWA-OV) & & & & \\ \\
    HERA & 70 & 19 & 154 & 150 \\
    (HERA-19) \\ \\
    HERA & 98 & 37 & 154 & 150 \\
    (HERA-37) \\ \\
    HERA & 294 & 331 & 154 & 150 \\
    (HERA-331) \\ \\
    HERA & 1330 & 6769\tnote{b} & 154 & 150 \\
    (HERA-6769) \\ \\
    SKA1-Low & 1000 & 750\tnote{c} & 962 & 150 \\
    (SKA1-LC) \\ \\
    SKA1-Low & 1000 & 192,000\tnote{c} & 962 & 150 \\
    Dipoles\tnote{d} \\
    (SKA1-LCD) \\
    \hline
  \end{tabular}
  \begin{tablenotes}
    \item[a] Owens Valley LWA
    \item[b] Hypothetically chosen to have a total collecting area of 
      1~km$^2$
    \item[c] This is the number of elements expected to be in the core, 
      three-fourths of the total number
    \item[d] All dipoles are used as independent elements without station 
      beamforming
  \end{tablenotes}
  \end{threeparttable}
\end{table}
\footnotetext{LWA at Ovens Valley}

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of 
complex operations per frequency channel per integration timescale. Telescopes 
that fall to the left of this line indicate MOFF imaging is computationally
more efficient than visibility based imaging. All HERA layouts except possibly
HERA-19 are in a parameter space where MOFF imaging holds the advantage. The 
solid line showing future trajectory of HERA like systems will be clearly 
favoured by MOFF imaging. The gray shaded area is for a projected LWA 
expansion and is also predominantly in the region favouring MOFF imaging. 
It is bounded by the LWA1 and LWA-OV on the left and right 
respectively. The current (see Table~\ref{tab:antenna-layouts} and
an expanded layout with a four-fold increase in number of elements over a
two-fold increase in $b_\textrm{max}$ provide the bounds at the bottom and top 
respectively. Current instruments such as MWA and LOFAR lie in parameter space
favouring visibility based imaging.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex multiplies and adds with MOFF and FX. The dashed line
    is the boundary at which the number of operations with MOFF and visibility 
    based imaging are equal. MOFF imaging is more efficient for telescopes 
    occupying the left of this line and vice versa. All the HERA layouts except
    possibly HERA-19 lie in the parameter space favoured by MOFF imaging. And so 
    are SKA1-LC and SKA1-LCD. The solid black curve shows the projected 
    trajectory of bigger close-packed hexagonal layouts similar to HERA. The 
    gray shaded area denotes the projected trajectory of the LWA bounded by LWA1 
    (left edge), LWA-OV (right edge), current layout (bottom) and a four-fold 
    increase in the number of elements in a two-fold increase in the core size
    (top). Current instruments such as MWA and LOFAR fall in a region favoured 
    by visibility based imaging.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

We now consider antenna array layouts described by three quantities essential
to radio interferometry, namely, maximum baseline length, number of antennas,
and the size of each antenna.

Fig.~\ref{fig:parameter-space-bll-nant-instruments} shows the ratio of the
number of computations required with visibility based imaging relative to 
MOFF imaging and the boundaries where this ratio is unity. The color scale 
refers to the ratio while using a 1~m antenna element. Pairs of black and
white lines of the same line style correspond to a specific antenna size. For 
a given antenna size (annotated on the plot), the white line denotes the 
minimum baseline length that could achieve the closest packing as a function 
of the number of antennas. Regions to the right of the white line for a 
given antenna size imply a physically impossible scenario where antennas will 
have to be packed overlapping with each other. For the same antenna size, the 
black line denotes the boundary where the ratio of the number of computations 
with either algorithm becomes unity. Regions to the right of the black lines 
favour usage of the MOFF algorithm. Thus the wedge between the white and black 
lines denotes the region in parameter space where MOFF algorithm holds a 
definite advantage while visibility based algorithms will be favoured to the 
left of the black lines. As antenna size increases the maximum number of 
antennas for a dense packing as a function of baseline length decreases. Hence 
the white lines shift leftward as antenna size increases. Similarly, with 
increase in antenna size, $N_\textrm{g}$ also decreases when field of view 
imaging is achieved with an increasing grid spacing equal to antenna size and
hence lowers the amount of computations required with the MOFF algorithm. This
shifts the black curves leftward. HERA-331, HERA-1027, LWA1, SKA1-LC and 
SKA1-LCD are inside their respective wedges indicating they are in regions that
favour the usage of MOFF imaging.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding.eps}
  \caption{Current and instruments planned for future in parameter space of
    baseline length and number of antennas with MOFF and FX. Color scale shown 
    is for the ratio of number of computations required by visibility based
    imaging to that by the MOFF algorithm. Different line styles denote 
    different antenna sizes annotated. For a given antenna size, the 
    corresponding white line denotes the maximum number of antennas that can 
    be packed inside various baseline lengths. The region to the right of the 
    white lines for corresponding antenna size is physically disallowed. 
    Regions to the left of the black lines of the same corresponding antenna 
    size (or line style) favours usage of visibility based imaging. Region 
    inside the wedge enclosed by the black and white lines favours using the 
    MOFF algorithm for any given antenna size. These wedges shift leftward with 
    increasing antenna size. HERA-331, HERA-1027, LWA1, SKA1-LC and SKA1-LCD
    are inside their corresponding wedges that favour the usage of MOFF 
    imaging.}
  \label{fig:parameter-space-bll-nant-instruments}
\end{figure}
